Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2024-05-19 doc/manuscript/supplementary_information.qmd
Revise manuscript for Nature Communications

---
#format: docx
format:
   latex: 
      cite-method: natbib
include-in-header:
  text: |
    \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
bibliography: translocation.bib
csl: pnas.csl
---

## Title: Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### Authors:

Roland A. Knapp^a,b,^\*, Mark Q. Wilber^c,1^, Allison Q. Byrne^d,e,1^, Maxwell B. Joseph^f,g^, Thomas C. Smith^a,b^, Andrew P. Rothstein^d,e,h^, Robert L. Grasso^i^, Erica Bree Rosenblum^d,e^

### Affiliations:

^a^Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546

^b^Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060

^c^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996

^d^Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114

^e^Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160

^f^Earth Lab, University of Colorado, Boulder, CO, 80303

^g^Current affiliation: Planet, San Francisco, CA, 94107

^h^Current affiliation: Ginkgo Bioworks, Boston, MA, 02210

^i^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318

\*Corresponding author: Sierra Nevada Aquatic Research Laboratory, 1016 Mount Morrison Road, Mammoth Lakes, CA 93546, phone: 760.647.0034, email: roland.knapp\@ucsb.edu

^1^These authors contributed equally to this work.

**Classification:** Biological Sciences - Ecology

**Key words:** amphibians, disease, evolution, reintroduction, population recovery

**Author contributions**: All authors designed research; R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., A.P.R., and R.L.G. performed research; M.B.J. contributed new analytic tools; R.A.K., M.Q.W., A.Q.B., A.P.R., and E.B.R. analyzed data; and R.A.K., M.Q.W., A.Q.B., M.B.J., T.C.S., and E.B.R. wrote paper.

**Competing interests:** This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).
The authors declare no additional competing interests.

### Abstract

Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in infectious diseases.
The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
Lacking any broadly applicable methods to reverse these impacts, the future of many amphibians appears grim.
The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes.
We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
In addition, the likelihood of establishment is influenced by site, population, and frog attributes.
Results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
We found numerous differences between these categories, including in immune function loci associated with disease resistance, consistent with evolutionary changes in response to Bd exposure.
These results provide a rare example of how reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species, and have broad implications for the many taxa worldwide that are threatened with extinction by novel pathogens.

### Significance

Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
We used reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to near-extinction by a novel disease that has decimated amphibians worldwide.
Reintroductions allowed establishment of viable populations in the presence of disease.
In addition, resistance may be at least partially the result of natural selection at specific immune function genes.
The evolution of resistance and reintroduction of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.

### Introduction

Human activities are increasingly impacting global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amphibians are experiencing particularly devastating impacts of disease due to the recent emergence and global spread of the highly virulent amphibian chytrid fungus, *Batrachochytrium dendrobatidis* (Bd) [@luedtke2023; @scheele2019].
By one estimate, hundreds of species have experienced Bd-caused declines, and numerous susceptible taxa are extinct in the wild [@scheele2019].
These impacts to global biodiversity may be unprecedented, and highlight the importance of understanding mechanisms of species persistence in the presence of emerging diseases [@russell2020].

Evidence of natural recovery in the many Bd-impacted amphibian populations is surprisingly rare [for notable exceptions, see @scheele2017; @voyles2018; @knapp2016], suggesting that disease-caused declines will be difficult to reverse.
This apparent low resilience to disease effects may be due to the limited ability of many amphibians to develop Bd resistance and/or tolerance, which in turn, could also lessen the effectiveness of potential Bd mitigation strategies.
Following pathogen arrival in a host population, resistance (ability to limit pathogen burden) and tolerance (ability to limit the harm caused by a particular burden) are key mechanisms to reduce disease impacts [@raberg2009] and facilitate population persistence and recovery [@brannelly2021].
Host immunity and evolution both play important roles in the development of resistance and tolerance, and an enhanced understanding of these processes is important for developing effective strategies to mitigate disease impacts in the wild [@garner2016; @woodhams2011].
However, several aspects of the amphibian-Bd system present difficult obstacles for the development of mitigation efforts, including (i) the general inability of amphibians to mount an effective immune response against Bd infection [@rosenblum2012; @fites2013; @grogan2018a], and (ii) the apparent rarity of evolution of more resistant/tolerant genotypes [but see @savage2016; @grogan2018b].
These factors suggest that reintroduction of amphibians into sites to reestablish populations extirpated by Bd will often result not in population recovery, but instead in the rapid reinfection and mortality of the introduced animals and/or their progeny [@hammond2021; @knapp2022; @stockwell2008; @scheele2021].
If true, the future of many amphibian species threatened by Bd appears bleak.

The mountain yellow-legged (MYL) frog, composed of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the global decline of amphibians caused by Bd [@scheele2019].
Once the most common amphibian in the high elevation portion of California's Sierra Nevada mountains [USA, @grinnell1924], during the past century this frog has disappeared from more than 90% of its historical range [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] starting in the late 1800s.
The arrival of Bd in the mid-1900s and its subsequent spread [@vredenburg2019] caused additional large-scale population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because fish eradication is feasible [@knapp1998] and results in the rapid recovery of frog populations [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [during which Bd infection intensity on frogs is very high, @briggs2010] and are now recovering (@fig-recovery-model) [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection [@knapp2016], with infection intensity ("load") on adults consistently in the low-to-moderate range [@briggs2010; @knapp2011; @joseph2018].
This reduced susceptibility is evident even under controlled laboratory conditions [@knapp2016], indicative of host resistance against Bd infection and not simply an effect of factors external to individual frogs (e.g., environmental conditions).
In addition to frogs from recovering populations having higher resistance to Bd infection than those from naive populations, they could also have higher tolerance, but no data are currently available to evaluate this possibility.
Therefore, we focus on resistance throughout this paper.
The observed resistance of MYL frogs could be the result of several mechanisms that are not mutually exclusive, including natural selection for more resistant genotypes [@savage2016; @grogan2018b], acquired immunity [@grogan2018a], and/or inherent between-population differences that pre-date Bd exposure.
The possible evolution of MYL frog resistance and subsequent population recovery is consistent with that expected under "evolutionary rescue", whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth [@carlson2014; @searle2020].
This intriguing possibility also suggests an opportunity to expand recovery beyond the spatial scale possible under natural recovery by using resistant frogs from recovering populations in reintroductions to vacant habitats (@fig-recovery-model) [@joseph2018; @mendelson2019].

In the current study, we had three primary objectives.
First, determine whether the reintroduction of resistant MYL frogs, obtained from populations recovering from Bd-caused declines, allows the reestablishment of populations despite ongoing disease (@fig-recovery-model).
We accomplished this using a 15-year large-scale reintroduction study.
Second, extend our inferences of population recovery beyond the temporal bounds of our reintroduction study to estimate the probability of persistence for the reintroduced populations over a multi-decadal period (@fig-recovery-model).
We did this using a stage-structured matrix model.
Third, given the importance of resistance for frog survival, population establishment, and long-term viability (this study), determine whether MYL frogs in recovering populations show genomic patterns consistent with selection for resistance (@fig-recovery-model).
We achieved this with a comparative study using exome capture methods applied to naive and recovering populations.

### Results

#### Frog population recovery

```{r}
#| include: false
#| code-fold: true

library(readr)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))

transno_min <- d1 %>% 
  distinct(site_id, date) %>% 
  group_by(site_id) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(min(n))

transno_max <- d1 %>% 
  distinct(site_id, date) %>% 
  group_by(site_id) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(max(n))

transfrogs_max <- d1 %>% 
  group_by(site_id, date) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(max(n))

transfrogs_min <- d1 %>% 
  group_by(site_id, date) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(min(n))

transfrogs_median <- d1 %>% 
  group_by(site_id, date) %>% 
  count() %>% 
  ungroup() %>% 
  summarize(median(n))

bdload_min <- d1 %>% 
  filter(!is.na(bd_load)) %>% 
  group_by(site_id) %>% 
  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
  ungroup() %>% 
  summarize(min(bdload_ave))

bdload_max <- d1 %>% 
  filter(!is.na(bd_load)) %>% 
  group_by(site_id) %>% 
  mutate(bdload_ave = median(log10(bd_load + 1))) %>% 
  ungroup() %>% 
  summarize(max(bdload_ave))

bdprev_max <- d1 %>% 
  filter(!is.na(bd_load)) %>% 
  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
  group_by(site_id) %>% 
  summarize(prev = sum(bdload01)/n()) %>% 
  ungroup() %>% select(prev) %>% 
  slice_max(prev, n = 1, with_ties = FALSE)

bdprev_min <- d1 %>% 
  filter(!is.na(bd_load)) %>% 
  mutate(bdload01 = if_else(bd_load > 0, 1, 0)) %>% 
  group_by(site_id) %>% 
  summarize(prev = sum(bdload01)/n()) %>% 
  ungroup() %>% select(prev) %>% 
  slice_min(prev, n = 1, with_ties = FALSE)

files <- fs::dir_ls(here::here("data", "raw", "cmr-analysis", "survey"))  
surveys <- read_csv(files, id = "path") %>% 
  filter(survey_date < "2023-01-01") %>% 
  select(-path)
rm(files)

surveys_yr_median <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr") %>% 
  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
  group_by(site_id) %>% 
  count(yr) %>% ungroup() %>% 
  summarize(median(n))

surveys_yr_max <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr") %>% 
  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
  group_by(site_id) %>% 
  count(yr) %>% ungroup() %>% 
  summarize(max(n))

surveys_yr_min <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr") %>% 
  distinct(site_id, primary_period, .keep_all = TRUE) %>% 
  group_by(site_id) %>% 
  count(yr) %>% ungroup() %>% 
  summarize(min(n))

robust_year_median <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr" & secondary_period > 1) %>% 
  group_by(site_id) %>% 
  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
  summarize(median(max_2period))

robust_year_min <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr" & secondary_period > 1) %>% 
  group_by(site_id) %>% 
  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
  summarize(min(max_2period))

robust_year_max <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr" & secondary_period > 1) %>% 
  group_by(site_id) %>% 
  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
  summarize(max(max_2period))

surveyyrs <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr") %>% 
  distinct(site_id, yr) %>% 
  count(site_id)

surveydates <- surveys %>% 
  filter(survey_type == "cmr") %>% 
  mutate(moday = format(as.Date(survey_date), "%b-%d"))

translocationdates <- surveys %>% 
  filter(survey_type == "translocation") %>% 
  mutate(moday = format(as.Date(survey_date), "%b-%d"))

frog_counts_gt0_nrow <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  nrow()

frog_counts_gt0_median <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(median(pyears_tadsub_present)) %>% 
  round(2)

frog_counts_gt0_min <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(min(pyears_tadsub_present)) %>% 
  round(2)

frog_counts_gt0_max <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(max(pyears_tadsub_present))

frog_captures_gt0_nrow <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  nrow()

frog_captures_gt0_median <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(median(pyears_recruits_present)) %>% 
  round(2)

frog_captures_gt0_min <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(min(pyears_recruits_present)) %>% 
  round(2)

frog_captures_gt0_max <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(max(pyears_recruits_present))
```

To determine whether MYL frogs from recovering populations can be used to reestablish extirpated populations, we conducted `r n_distinct(d1$site_id, d1$date)` reintroductions in Yosemite National Park (`r min(d1$year)`--`r max(d1$year)`).
Each of the reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites (@fig-yosemap SI).
The 3 donor populations are among the largest MYL frog populations in Yosemite, and included 2 of the 5 recovering populations used in the frog evolution study referenced above (see **Results - Frog genome evolution:** populations 1 and 4).
These 2 donor populations provided frogs for 20 of the `r n_distinct(d1$site_id, d1$date)` translocations.
Following translocation, we estimated adult survival and recruitment of new adults from capture-mark-recapture (CMR) surveys and obtained counts of tadpoles and juveniles from visual encounter surveys (VES).
Across all translocation sites, the duration of survey time series was `r min(surveyyrs$n)`--`r max(surveyyrs$n)` years (median = `r (median(surveyyrs$n))`).

Of the `r nrow(d2)` reintroduced populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.

Bd loads were fairly consistent before versus after translocation, and loads were nearly always well below the level indicative of severe chytridiomycosis (i.e., the disease caused by Bd) and associated frog mortality (@fig-bdload-beforeafter SI) [@joseph2018; @vredenburg2010].
Although it is possible that the observed relatively small changes in load are a consequence of individuals with high Bd loads dying and therefore being unavailable for sampling during the post-translocation period, the fact that there was little difference in pre- versus post-translocation Bd loads even in those populations that had very high frog survival (70556, 74976 - see below; @fig-bdload-beforeafter SI) suggests a true lack of substantial change in Bd load.

The ultimate measure of reintroduction success is the establishment of a self-sustaining population.
Given that it can take years or even decades to determine the self-sustainability of a reintroduced population [for an example in MYL frogs, see @joseph2018], the use of proxies is essential for providing shorter-term insights into reintroduction success and the factors driving it.
Results from our CMR surveys allowed us to accurately estimate frog survival, including over the entire CMR time series for each site and during only the 1-year period immediately following translocation.
These estimates were made using site-specific models analyzed using the mrmr package.
We use these estimates to describe general patterns of frog survival in all translocated cohorts, and in an among-site meta-analysis of frog survival to identify important predictors of 1-year frog survival (e.g., Bd load).

Estimates of 1-year frog survival indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
These patterns indicate an important effect of site characteristics on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier: 5 of the 7 populations translocated after 2013 had estimated survival $\ge$ 0.5, compared to only 1 of 5 populations translocated prior to 2013.
We suggest this resulted primarily from our improved ability to choose recipient sites with higher habitat quality for *R. sierrae* (see **Materials and Methods - Frog population recovery - Field methods** for details).
This increased survival has direct implications for population viability (see **Results - Long-term population viability**).

The goal of our meta-analysis was to identify important predictors of 1-year frog survival.
We were particularly interested in whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
This analysis was conducted in a Bayesian framework and included a diversity of site, cohort, and individual-level characteristics as predictors and 1-year frog survival (@fig-translocation-survival) as the response variable.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (@fig-survival-postdens SI).
Instead, important predictors included winter severity in the year following translocation (snow_t1), site elevation, and donor population (@fig-cond-effects, @fig-survival-postdens SI).
Males had somewhat higher survival than females, but this effect was small (@fig-cond-effects, @fig-survival-postdens SI).
The absence of Bd load as an important predictor of frog survival is consistent with frogs in recovering populations having sufficient resistance to suppress Bd loads below harmful levels.

In summary, results from our frog translocation study indicate that (i) translocations produced relatively high 1-year survival of translocated adults, as well as reproduction and recruitment, at the majority of recipient sites, (ii) 1-year survival of adults is influenced by site characteristics, weather conditions, and donor population (but not Bd load), and (iii) based on the relatively small changes in Bd load after translocation, loads appear more strongly influenced by frog characteristics (e.g., resistance) than site characteristics.
Together, these results indicate that frogs translocated from recovering populations can maintain the benefits of resistance in non-natal habitats.
In addition, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally-recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
This suggests that frog resistance is maintained across generations.
All of the conditions described above are supportive of population establishment and long-term population growth.

#### Long-term population viability

Results from the frog translocation study are suggestive of population establishment.
However, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient for long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).

Given observed yearly adult survival probabilities of translocated frogs (from site-specific mrmr CMR models; provided in legend of @fig-viability B) and a yearly survival probability of the year-1 juvenile class ($\sigma_{J_1}$) greater than 0.09, at least six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.19-1.40 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
As year-1 juvenile survival probability increased above 0.2, the deterministic long-run growth rate of eight of twelve population was greater than 1 (@fig-viability A).

Even when incorporating (i) demographic stochasticity and (ii) environmental stochasticity in year-1 juvenile survival and recruitment (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities ($\sigma_{A_R} > 0.5$) had 50-year extinction probabilities of less than 0.5 when the average year-1 juvenile survival was greater than 0.10 (@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in survival and recruitment.
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be \> 50% regardless of the value of mean year-1 juvenile survival between 0 and 0.25.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.

#### Frog genome evolution

Results from the preceding sections indicate the critical role of intrinsic factors (such as frog resistance) in post-translocation frog survival, population growth, and population viability.
As such, identifying the mechanisms underlying resistance would fill a key gap in our understanding of the factors that promote population resilience in the presence of disease.
Although natural selection for more resistant frog genotypes, and evolutionary rescue, may be foundational to the ability of frogs to recover despite ongoing Bd infection, for MYL frogs the role of disease-mediated selection in these processes remains unknown.

To determine whether MYL frog populations show genomic patterns consistent with an evolutionary response to Bd, we compared frog exomes (i.e., coding region of a genome) between populations with contrasting histories of Bd exposure.
Specifically, we compared frog genomes sampled in 4 populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in 5 populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
Bd-exposure histories of the 9 study populations are based on 10-20 years of VES and Bd surveillance using skin swabbing [e.g., @knapp2016; @zhou2015; @wilber2022].
Naive populations are characterized by large numbers of adults (i.e, typically 1000s), Bd prevalence that is generally 0% except during occasional Bd failed invasions [during which Bd loads remain very low, @wilber2022], and no history of Bd epizootics since we first surveyed these populations in the late 1990s and early 2000s [@zhou2015].
In contrast, recovering populations exist in an enzootic state [@briggs2010], characterized by smaller numbers of adults (generally \< 500), high Bd prevalence [often \> 80%, @knapp2011], and, in adults, moderate Bd loads that are typically well below the level expected to cause mortality [@vredenburg2010].
Naive and recovering populations can be identified unambiguously using these differences in Bd prevalence and load.
There is no potential for frog dispersal between the 9 study populations or any other populations due to intervening distances and topography, as well as the presence of introduced (predatory) fish and fish-induced habitat fragmentation.

It is important to note that our genomic analysis is necessarily limited by the small number of naive populations remaining on the landscape (nearly all of which we sampled as part of the current study) and their clumped distribution ([@fig-allelefrequencies]).
These characteristics of naive populations precluded a replicated analysis among pairs of geographically proximate naive and recovering populations.
Instead, to obtain adequate power, we conducted a pooled analysis across population types to identify genomic regions showing patterns consistent with selection.
Our comparative approach cannot provide a direct assessment of the strength or timing of natural selection, but does allow us to identify specific regions of the genome that merit further study for their potential contribution to frog resistance in recovering populations.
(See **Supporting Information - Frog genome evolution - Study design** for additional details.)

As a first step in our pooled analysis, we conducted a principal component analysis (PCA) of the genomic data to describe the relationships between sampled populations.
We then used two complementary approaches to identify regions of the genome that differed between naive and recovering populations (i.e., regions possibly under selection).
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants, including single nucleotide polymorphisms (SNPs) and insertions/deletions (INDELS), while accounting for population structure.
Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).

Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A SI), and clusters reflected the species split (i.e., *R. muscosa* versus *R. sierrae*) and the strong signature of isolation-by-distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017; @byrne2023].
Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish allelic associations of individuals sampled in the 2 population types versus allelic associations resulting from population structure and genetic drift.

Results from the individual variant and splined window analyses show that recovering populations show patterns consistent with selection on multiple regions of the genome.
The analysis of individual variants identified 11 "outlier" SNPs (i.e., showing significantly different allele frequencies between naive versus recovering populations) from 7 distinct genes across 4 contigs (@fig-selectionresults B, C).
One of the 7 identified genes (LOC108802036) does not have an associated annotation.
For the outlier SNPs, frequency differences between the naive and recovering populations ranged from 0.41 to 0.86.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of parallel selection at this locus across multiple populations.
The other 6 outlier variants showed less consistent frequency differences across the study populations, but for these we still found statistically significant differences in 2 of the 3 genetic clusters (containing populations 5--9; @fig-allelefrequencies, @fig-selectionresults A SI).
Therefore, although some outlier variant associations have a more limited geographic extent than RIN3, they still describe results that are consistent with parallel evolutionary changes.

The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics ("shared regions") and 2 of these shared regions also contained one or more of the outlier SNPs described above.
A total of 35 annotated genes were found in the 9 shared regions.
Given this large number of genes, here we focus on those with the strongest signal of selection and/or immune-related functions.
The largest $\pi_{diff}$, suggestive of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had a large negative $\pi_{diff}$ value, consistent with balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot SI).
This region (854kb from the beginning of the outlier region to the VARS SNP) contained a total of 8 annotated genes.
In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, this region is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.
To provide details specifically on possible selection in the Yosemite populations described in **Results - Frog population recovery**, $\pi$ values for the 9 shared outlier regions are shown in @fig-ave-pi-per-popn-yose SI for each of the four Yosemite populations; two of these populations provided frogs for 20 of the 24 frog translocations.

Although the joint processes of Bd-caused population declines and selection in response to Bd exposure could affect genetic diversity of recovering populations, we found no consistent differences in individual-level heterozygosity or population-level $\pi$ between naive and recovering populations (see **Supporting Information - Frog genomic evolution - Genetic diversity** for details).
Thus, despite patterns consistent with localized selection in particular regions of the genome, we did not find evidence for reduced genetic diversity across the genome in recovering populations.
In addition, no gene ontology (GO) biological functions, molecular functions, or cellular processes were over-represented in either the outlier variants or the 35 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows (see **Supporting Information - Frog genomic evolution** **- GO analysis** for details).

In summary, our genomic results indicate that frogs from naive and recovering populations show differences at several immune-function loci, consistent with selection following Bd exposure.

### Discussion

Disease-induced population declines are decimating global biodiversity [@daszak2000], but broadly-applicable strategies to recover affected species are generally lacking [e.g., @garner2016].
Here, we tested the possibility that populations of resistant individuals from naturally recovering populations can be used to reestablish extirpated populations of the endangered MYL frog in the presence of a highly virulent pathogen (Bd).
Our results indicate (i) the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, (ii) that 50% of the reintroduced populations are likely to persist over a 50-year period, (iii) that there are genomic differences between naive and recovering MYL frog populations, consistent with evolutionary change in frogs following Bd exposure, and (iv) that some of the genomic differences are in regions that contain genes related to disease resistance.
Collectively, these results (@fig-recovery-model) provide a rare example of amphibian recovery in the presence of Bd, and have important implications for the conservation and recovery of amphibians and other taxa worldwide that are endangered by escalating impacts from emerging infectious diseases.
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via reintroduction of resistant individuals is striking, and even more so given that MYL frogs were driven to near-extinction by Bd.

In the following discussion, we follow the sequence of frog recovery described in @fig-recovery-model to structure our key points.
Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter SI].
The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [see Figure 4 in @knapp2016], and clearly indicate that frogs from recovering populations exhibit resistance against Bd infection.
This resistance could be due at least in part to natural selection for more resistant genotypes, but until now evidence to evaluate the possible role of evolution was lacking.

Our genomics study used a pooled design that compared MYL frogs from naive and recovering populations distributed across a similar latitudinal range.
The study design produced frog genetic clusters that each contained at least one naive and one recovering population, allowing us to infer selection while attempting to minimize the confounding effects of population structure.
We identified multiple specific alleles and genomic regions showing differences between naive and recovering MYL frog populations, consistent with selection following Bd exposure.
In addition, we did not find a reduction in overall genetic variation in recovering populations, suggesting that despite possible localized selection in the genome, some of these populations likely retain adequate genetic diversity for long-term persistence.

>>>>

---
format: pdf
#format:
#   latex: 
#      cite-method: natbib
include-in-header:
  text: |
    \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
crossref: 
  fig-title: "**Supplementary Fig.**"
  tbl-title: "**Supplementary Table**"
  title-delim: "|"
bibliography: translocation.bib
csl: nature.csl
---

### Supplementary Information for

## Reintroduction of resistant frogs facilitates recovery in the presence of a lethal fungal disease

Roland A. Knapp\*, Mark Q. Wilber, Maxwell B. Joseph, Thomas C. Smith, Robert L. Grasso

\* **Author for correspondence:** roland.knapp\@ucsb.edu
